LCS Tool : A Computational Platform for Lagrangian Coherent Structures

### 2014-08-23 main.tex
Correct spelling and grammar in main.tex
Commit da33cca incorporated changes made in the arXiv e-print
including mistakes. The present commit corrects those mistakes.

Lagrangian Coherent Structures (LCSs) are evolving organizing centers of trajectory patterns in non-autonomous dynamical systems\parencite{haller00:_lagran,peacock13:_lagran,Haller2014}. Applications of LCSs include oceanic and atmospheric flows\parencite{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\parencite{gawlik09:_lagran}, crowd dynamics\parencite{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\parencite{hadjighasem13:_detec_kam}.

\textcite{haller01:_distin} proposed that ridges of the Finite-Time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling- and attracting-type) LCSs. A number of examples indeed support this principle\parencite{peacock13:_lagran}. Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \parencite{beron-vera10:_invar_lagran}).

@@

The right Cauchy-Green strain tensor associated with the flow map is defined as
\begin{equation}
C_{t_0}^t(x_0) = \left[\nabla F_{t_0}^t(x_0)\right]^T \nabla F_{t_0}^t(x_0),
\label{eq:CG}
\end{equation}
measuring Lagrangian strain in the velocity field. This tensor is symmetric and positive definite\parencite{truesdell04}. We label the eigenvalues and eigenvectors of $C_{t_0}^t(\boldsymbol x_0)$ as follows:
\[
C_{t_0}^t \xi_i = \lambda_i \xi_i, \qquad 0 < \lambda_1 \leq \lambda_2, \quad i = 1,2;
\]
\begin{equation}
\left|\xi_i\right| = 1, \quad \xi_2 = \Omega \xi_1, \quad \Omega = \left(
\begin{array}{rr}
0 & -1\\
1 & 0
\end{array}
\right).
\label{eq:CG_invariants}
\end{equation}

@@

The auxiliary grid (cf. Table~\ref{t:Cauchy-Green algorithm}) comprises four points placed symmetrically around each point of the Cartesian grid (Figure~\ref{f:main and auxiliary grids}.) These points are used to achieve increased accuracy in the finite-difference approximation
\[
\nabla F_{t_0}^t(x_0) \approx
\left(\begin{array}{cc}
\frac{x_1(t;t_0,x_0 + \delta x_1) - x_1(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_1(t;t_0,x_0+\delta x_2) - x_1(t;t_0,x_0-\delta x_2)}{2\vert\delta x_2\vert}\\
\frac{x_2(t;t_0,x_0+\delta x_1) - x^2(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_2(t;t_0,x_0 + \delta x_2) - x_2(t;t_0,x_0 - \delta x_2)}{2\vert\delta x_2\vert}
\end{array}\right)
\]
of $\nabla F_{t_0}^t$ at a point $x_0$ of the Cartesian grid. Here $\delta x_i$ is a vector of length $\vert\delta x_i\vert > 0$ that points from the Cartesian grid-point $x_0$ in the $i^{th}$ coordinate direction (Figure~\ref{f:main and auxiliary grids}). Computational improvements arising from the use of the auxiliary grid over simply using the nearest points of the main grid were reported in Ref.\,\parencite{farazmand12:_comput_lagran}. Experience suggests an auxiliary grid spacing of 1-10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=.5\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of the main grid (filled circles) and the auxiliary grid (empty circles) used in the computation of the derivative of the flow map in the definition \eqref{eq:CG} of the Cauchy--Green stain tensor. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

The function \lstinline!eig_cgStrain! of LCS Tool provides the option to calculate Cauchy--Green eigenvectors from the auxiliary grid using eigenvalues calculated from the main grid. We have found that for flows defined analytically, the eigenvalues can reliably be calculated from the main grid. For the flows defined by datasets, using the auxiliary grid for eigenvalue calculations gives better results.

As mentioned above, a typical main grid for the Cauchy-Green strain tensor comprises $500 \times 500$. This means that after the addition of 4 auxiliary grid points around each main grid points, eq. \eqref{eq:v(x,t)} must be integrated over 1.25 million initial conditions.

To avoid excessive computational times in MATLAB, we vectorize \eqref{eq:v(x,t)}, i.e., combine its right-hand-side evaluated over each initial points into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ODEs. We then use MATLAB's \lstinline!ode45! to perform trajectory integration from all grid points simultaneously. This process typically takes five to ten minutes with default error tolerances. A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

The Cauchy-Green strain tensor generically admits points singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined. This generically arises at a finite set of isolated points within the computational domain\parencite{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default values: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default value: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{Syntax of LCS Tool's elliptic LCS functions.}
\label{t:Elliptic LCS functions}
\end{table}

@@

As discussed in section \ref{sub:Hyperbolic-LCSs}, positions of hyperbolic LCSs at time $t_{0}$ are found as the strongest repelling orbits of the vector field \eqref{eq:strainline} (repelling LCSs), and strongest attracting orbits of the vector field \eqref{eq:stretchline} (attracting LCSs.) By repulsion and attraction we mean a property of the LCS (as an evolving material line) under the flow map $F_{t_0}^t$. We identify the strongest repelling shrinklines as those crossing a local maximum of the $\lambda_2(x_0)$ field. Similarly, we identify the strongest attracting stretchlines as those crossing a local minimum of the $\lambda_1(x_0)$ field. These local maxima and minima of the appropriate $\lambda_i(x_0)$ eigenvalue field can be thought of as the extensions of the concept of a saddle points to the present finite-time, temporally aperiodic flow setting.

@@

\subsection{Double gyre}
The double gyre is a model for a time-dependent two gyre system observed in geophysical flows \parencite{shadden05:_defin_lagran_lyapun}. The model consists of two counter rotating sinusoidal vortices with a harmonically oscillating line in-between. Lagrangian particle motions satisfy the non-autonomous dynamical system
\begin{equation}
\begin{split}
\od{x}{t} = -\pi A \sin[\pi f(x,t)] \cos(\pi y),\\
\od{y}{t} = \pi A \cos[\pi f(x,t)] \sin(\pi y) \pd{f(x,t)}{x},\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x.
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function describing this model velocity field is given in Listing~\ref{l:double gyre derivative}, specifying the right hand side of the particle ODE in a way that supports vectorized integration.

@@

In Figure~\ref{fig:double_gyre_lambda_lcs_convergence}, the resolution of the  Cauchy-Green strain tensor is varied from $500 \times 250$ to $1000 \times 500$. The location of the outermost closed lambda-line changes little, demonstrating convergence. For all tested resolutions the $\lambda$ values of the outermost closed orbits are constant. This is also suggests that the lowest tested resolution of $500 \times 250$ is sufficient to identify the elliptic LCS in this example.

@@

This last example demonstrates the use of LCS Tool on velocity data derived from satellite-observed sea-surface heights under the geostrophic approximation. In contrast to the previous two analytic examples, the velocity field is available only with discrete temporal and spatial resolution. Our region of interest is a small domain in the South Atlantic Ocean, where exceptionally coherent eddies, Agulhas rings, were recently found by \parencite{haller13:_coher_lagran} using the theory we surveyed in Section \ref{sub:Elliptic-LCSs}.

@@

Figures~\ref{f:ocean dataset hyperbolic shear lcs details strainline} and \ref{f:ocean dataset hyperbolic shear lcs details stretchline} show the positions of elliptic and hyperbolic LCSs on 22 November 2006, which is the same time instant analyzed in~\textcite{haller13:_coher_lagran,beron-vera13:_objec_agulh}. Here the integration time is chosen as $T = 30$ days, as opposed to 90 days in the cited references, to avoid the length tangling of hyperbolic LCSs. Our analysis via LCS Tool reveals five coherent eddy boundaries. The large coherent eddy at $(3,-32)$ has a non-stretching boundary, i.e., $\lambda = 1$. It corresponds to eddy \#2 in Figure~3 of \textcite{beron-vera13:_objec_agulh}. Additionally, four further smaller coherent eddy cores are found. They do not stay coherent over a time of 90 days. All closed orbits are stable under increased spatial resolution for the Cauchy-Green strain tensor field. Repelling LCSs (red) and attracting LCSs (blue) are superimposed over the plot of FTLE values. The hyperbolic LCSs determine the deformation of the fluid in-between coherent eddy cores.

>>>>

Lagrangian Coherent Structures (LCSs) are evolving organizing centers of trajectory patterns in non-autonomous dynamical systems\parencite{haller00:_lagran,peacock13:_lagran,haller15:_langr_coher_struc}. Applications of LCSs include oceanic and atmospheric flows\parencite{beron-vera13:_objec_agulh,koh02:_hyper}, biological transport problems\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran,huhn12:_south_indian_ocean_count_madag}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor}, celestial mechanics\parencite{gawlik09:_lagran}, crowd dynamics\parencite{ali07:_lagran_partic_dynam_approac_crowd}, and aperiodically forced mechanical oscillators\parencite{hadjighasem13:_detec_kam}.

\textcite{haller01:_distin} proposed that ridges of the finite-time Lyapunov exponent (FTLE) are heuristic indicators of hyperbolic (i.e., repelling- and attracting-type) LCSs. A number of examples indeed support this principle\parencite{peacock13:_lagran}. Equating FTLE ridges with LCSs, however, would create theoretical inconsistencies, as well as false positives and negatives in hyperbolic LCS detection\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. In addition, the role of the FTLE field in the accurate detection of elliptic (vortex-type) and parabolic (jet-core type) LCSs has remained an open question (but see \parencite{beron-vera10:_invar_lagran}).

@@

The right Cauchy-Green strain tensor associated with the flow map is defined as
\begin{equation}
C_{t_0}^t(x_0) = \left[\nabla F_{t_0}^t(x_0)\right]^T \nabla F_{t_0}^t(x_0),
\label{eq:CG}
\end{equation}
measuring Lagrangian strain in the velocity field. This tensor is symmetric and positive definite\parencite{truesdell04}. We label the eigenvalues and eigenvectors of $C_{t_0}^t(x_0)$ as follows:
\[
C_{t_0}^t \xi_i = \lambda_i \xi_i, \qquad 0 < \lambda_1 \leq \lambda_2, \quad i = 1,2;
\]
\begin{equation}
\left|\xi_i\right| = 1, \quad \xi_2 = \Omega \xi_1, \quad \Omega = \left(
\begin{array}{rr}
0 & -1\\
1 & 0
\end{array}
\right).
\label{eq:CG_invariants}
\end{equation}

@@

The auxiliary grid (cf. Table~\ref{t:Cauchy-Green algorithm}) comprises four points placed symmetrically around each point of the Cartesian grid (Figure~\ref{f:main and auxiliary grids}.) These points are used to achieve increased accuracy in the finite-difference approximation
\[
\nabla F_{t_0}^t(x_0) \approx
\left(\begin{array}{cc}
\frac{x_1(t;t_0,x_0 + \delta x_1) - x_1(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_1(t;t_0,x_0+\delta x_2) - x_1(t;t_0,x_0-\delta x_2)}{2\vert\delta x_2\vert}\\
\frac{x_2(t;t_0,x_0+\delta x_1) - x_2(t;t_0,x_0 - \delta x_1)}{2\vert\delta x_1\vert} & \frac{x_2(t;t_0,x_0 + \delta x_2) - x_2(t;t_0,x_0 - \delta x_2)}{2\vert\delta x_2\vert}
\end{array}\right)
\]
of $\nabla F_{t_0}^t$ at a point $x_0$ of the Cartesian grid. Here $\delta x_i$ is a vector of length $\vert\delta x_i\vert > 0$ that points from the Cartesian grid-point $x_0$ in the $i^{th}$ coordinate direction (Figure~\ref{f:main and auxiliary grids}). Computational improvements arising from the use of the auxiliary grid over simply using the nearest points of the main grid were reported in Ref.\,\parencite{farazmand12:_comput_lagran}. Experience suggests an auxiliary grid spacing of 1-10\% of the main grid spacing.

\begin{figure}
\begin{center}
\includegraphics[width=.5\textwidth]{graphics/main_aux_grids}
\end{center}
\caption{Illustration of the main grid (filled circles) and the auxiliary grid (empty circles) used in the computation of the derivative of the flow map in definition \eqref{eq:CG} of the Cauchy-Green strain tensor. The variable \lstinline!auxGridRelDelta! specifies the grid spacing of the auxiliary grid relative to the main grid spacing.}
\label{f:main and auxiliary grids}
\end{figure}

The function \lstinline!eig_cgStrain! of LCS Tool provides the option to calculate Cauchy-Green eigenvectors from the auxiliary grid using eigenvalues calculated from the main grid. We have found that for flows defined analytically, the eigenvalues can reliably be calculated from the main grid. For flows defined by datasets, using the auxiliary grid for eigenvalue calculations gives better results.

As mentioned above, a typical main grid for the Cauchy-Green strain tensor comprises $500 \times 500$ points. This means that after the addition of 4 auxiliary grid points around each main grid point, eq. \eqref{eq:v(x,t)} must be integrated over 1.25 million initial conditions.

To avoid excessive computational times in MATLAB, we vectorize \eqref{eq:v(x,t)}, i.e., combine its right-hand-side evaluated over each initial point into a single system of equations. The resulting system is composed of independent blocks of two-dimensional first-order ODEs. We then use MATLAB's \lstinline!ode45! to perform trajectory integration from all grid points simultaneously. This process typically takes five to ten minutes with default error tolerances. A potential drawback of vector form integration is that memory requirements may become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two-dimensional form.

The Cauchy-Green strain tensor generically admits point singularities, i.e., points where $C_{t_0}^t(x_0)$ has repeated eigenvalues. At these points the eigenvectors $\xi_1(x_0)$ and $\xi_2(x_0)$ are no longer well-defined. This generically arises at a finite set of isolated points within the computational domain\parencite{delmarcelle94}, and hence lie off the computational grid with probability one.

@@

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline
\multicolumn{2}{|p{\textwidth}|}
{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\tabularnewline
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\tabularnewline
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\tabularnewline
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\tabularnewline
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\tabularnewline
\hline
\lstinline!domain! & array to define flow domain\tabularnewline
\hline
\lstinline!resolution! & $1 \times 2$ array to define main grid resolution for Cauchy-Green strain tensor\tabularnewline
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\tabularnewline
\hline
\lstinline!PSList! & user-defined structure for Poincare section end-points, number of $\lambda$-lines launched from Poincare section, and maximum closed $\lambda$-line length\tabularnewline
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare map. Default value: 5.\tabularnewline
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare map. Default value: $10^{-2}$.\tabularnewline
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify $\lambda$-line integration parameters\tabularnewline
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\tabularnewline
\hline
\end{tabular}
\caption{Syntax of LCS Tool's elliptic LCS functions.}
\label{t:Elliptic LCS functions}
\end{table}

@@

As discussed in section \ref{sub:Hyperbolic-LCSs}, positions of hyperbolic LCSs at time $t_{0}$ are found as the strongest repelling orbits of the vector field \eqref{eq:strainline} (repelling LCSs), and strongest attracting orbits of the vector field \eqref{eq:stretchline} (attracting LCSs.) By repulsion and attraction we mean a property of the LCS (as an evolving material line) under the flow map $F_{t_0}^t$. We identify the strongest repelling shrinklines as those crossing a local maximum of the $\lambda_2(x_0)$ field. Similarly, we identify the strongest attracting stretchlines as those crossing a local minimum of the $\lambda_1(x_0)$ field. These local maxima and minima of the appropriate $\lambda_i(x_0)$ eigenvalue field can be thought of as the extensions of the concept of saddle points to the present finite-time, temporally aperiodic flow setting.

@@

\subsection{Double gyre}
The double gyre is a model for a time-dependent two gyre system observed in geophysical flows \parencite{shadden05:_defin_lagran_lyapun}. The model consists of two counter rotating sinusoidal vortices with a harmonically oscillating line in-between. Lagrangian particle motions satisfy the non-autonomous dynamical system
\begin{equation}
\begin{split}
\od{x}{t} = -\pi A \sin[\pi f(x,t)] \cos(\pi y),\\
\od{y}{t} = \pi A \cos[\pi f(x,t)] \sin(\pi y) \pd{f(x,t)}{x},\\
f(x,t) = \epsilon \sin(\omega t) x^2 + [1 - 2 \epsilon \sin(\omega t)] x.
\end{split}
\label{eq:double gyre derivative equations}
\end{equation}
The MATLAB function describing this velocity field is given in Listing~\ref{l:double gyre derivative}, specifying the right hand side of the particle ODE in a way that supports vectorized integration.

@@

In Figure~\ref{fig:double_gyre_lambda_lcs_convergence}, the resolution of the  Cauchy-Green strain tensor is varied from $500 \times 250$ to $1000 \times 500$. The location of the outermost closed lambda-line changes little, demonstrating convergence. For all tested resolutions the $\lambda$ values of the outermost closed orbits are constant. This is also suggests that the lowest tested resolution of $500 \times 250$ is sufficient to identify the elliptic LCSs in this example.

@@

The final example demonstrates the use of LCS Tool on velocity data derived from satellite-observed sea-surface heights under the geostrophic approximation. In contrast to the previous two analytic examples, the velocity field is available only with discrete temporal and spatial resolution. Our region of interest is a small domain in the South Atlantic Ocean, where exceptionally coherent eddies, Agulhas rings, were recently found by \parencite{haller13:_coher_lagran} using the theory we surveyed in Section \ref{sub:Elliptic-LCSs}.

@@

Figures~\ref{f:ocean dataset hyperbolic shear lcs details strainline} and \ref{f:ocean dataset hyperbolic shear lcs details stretchline} show the positions of elliptic and hyperbolic LCSs on 22 November 2006, which is the same time instant analyzed in~\textcite{haller13:_coher_lagran,beron-vera13:_objec_agulh}. Here the integration time is chosen as $T = 30$ days, as opposed to 90 days in the cited references, to avoid tangling of hyperbolic LCSs. Our analysis via LCS Tool reveals five coherent eddy boundaries. The large coherent eddy at $(3,-32)$ has a non-stretching boundary, i.e., $\lambda = 1$. It corresponds to eddy \#2 in Figure~3 of \textcite{beron-vera13:_objec_agulh}. Additionally, four smaller coherent eddy cores are found. They do not stay coherent over a time of 90 days. All closed orbits are stable under increased spatial resolution for the Cauchy-Green strain tensor field. Repelling LCSs (red) and attracting LCSs (blue) are superimposed over the plot of FTLE values. The hyperbolic LCSs determine the deformation of the fluid in-between coherent eddy cores.
